### 2025-07-28 docs/article/Sections/results.tex
New Section 6 - Discussion

Virtual Threads allocate a full call stack per thread, which is unmounted and
stored on the heap when suspended (e.g., during blocking I/O). In contrast,
Kotlin coroutines compile into state machines that capture only the minimal
execution state needed to resume computation. This typically allows coroutines
to scale more efficiently in terms of memory usage, especially under high
concurrency. These findings contrast with prior research, such as Beronić et
al.~\cite{9803765}, which reported lower heap usage for Java Virtual Threads (16--64~MB)
compared to Kotlin coroutines (52--99~MB), albeit under different workloads.

These observations suggest that the choice between Virtual Threads and
suspendable coroutines should be guided by both runtime behavior and
development preferences. Virtual Threads offer the simplicity of synchronous
programming with competitive performance in moderate-load scenarios. In
contrast, suspendable coroutines, while requiring an asynchronous programming
model, may offer better efficiency in applications with higher concurrency,
deeper stacks, or more dynamic workloads.

>>>>

Virtual Threads allocate a full call stack per thread, which is unmounted and
stored on the heap when suspended (e.g., during blocking I/O). In contrast,
Kotlin coroutines compile into state machines that capture only the minimal
execution state needed to resume computation. This typically allows coroutines
to scale more efficiently in terms of memory usage, especially under high
concurrency.

\section{Discussion}

Beronić et al.~\cite{9803765} compared different structured concurrency constructs 
in Java and Kotlin in the context of a multithreaded HTTP server. 
Their benchmark scenario differs from ours: their server awaited incoming requests 
and, upon receiving one, executed a task involving object creation, writing the 
object's information to a file, and returning the data as a response.
Consistent with our findings, they concluded that both Kotlin's coroutines and 
Java's virtual threads offer significant performance improvements over traditional 
JVM-based threads in concurrent applications.

However, our results diverge from theirs regarding memory usage. 
They reported lower heap usage for Java virtual threads (16–64~MB) compared to 
Kotlin coroutines (52–99~MB), which contrasts with our measurements. 
This discrepancy may be attributed to the different workloads: while their 
benchmark includes a single I/O read-write operation per request, our experiments 
focus exclusively on read-only I/O tasks.

The work of Navarro et al.~\cite{navarro2023considerations} focused on the
performance of Quarkus, which relies on Eclipse Vert.x~\cite{vertx}, itself
built on top of Netty~\cite{netty}. 
This represents a similar environment to the one evaluated in our experiments
with Quarkus. 
Their study also emphasized template rendering, using a blocking HTML template
engine (Qute). 
The benchmark they employed is the Fortunes
test\footnote{\url{https://www.techempower.com/benchmarks}}, which involves
rendering a simple HTML table with only two data bindings per row.

Their results show that Quarkus with Virtual Threads outperformed the
traditional thread-per-request model under increased concurrency. Moreover,
compared to the reactive model, virtual threads demonstrated competitive
performance, particularly when executing blocking operations within template
engines.
In contrast, our experiments revealed that the reactive approach underperforms 
relative to virtual threads, likely due to the significantly larger number of 
data bindings in our use cases—namely, the \texttt{Presentation} and \texttt{Stock} 
models.

The work of Šimatović et al.\cite{vsimatovic2025evaluating} presents an
evaluation of Java Virtual Threads under different garbage collectors. They
explored three types of workloads, only one of which was based on a web
application scenario. This scenario replicated large-scale web scraping by
executing 1,000 parallel tasks, each consisting of an I/O-bound operation
followed by CPU-bound string processing.

In this workload, garbage collection activity remained low across all 
collectors, indicating reduced memory pressure. These findings suggest 
that Virtual Threads improve memory allocation efficiency and reduce the 
frequency of garbage collection cycles—particularly when used with 
concurrent garbage collectors.

Our observations suggest that the choice between Virtual Threads and
suspendable coroutines should be guided by both runtime behavior and
development preferences. Virtual Threads offer the simplicity of synchronous
programming with competitive performance in moderate-load scenarios. In
contrast, suspendable coroutines, while requiring an asynchronous programming
model, may offer better efficiency in applications with higher concurrency,
deeper stacks, or more dynamic workloads.
